Gradual Soundness: Lessons from Static Python

### 2021-10-12 ecoop-2022/paper.tex
ecoop: quick intro pass

\acknowledgements{
  Thanks to
   Guido van Rossom for provocative tweets.
  This work was partly supported by the US National Science Foundation.
  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
}

@@

Gradual typing has attracted significant interest as a solution to
the old tradeoff between static and dynamic typing.
The premise is simple: let programmers apply types to part of the
codebase and leave the rest un-analyzed.
Run-time checks can in principle enforce the assumptions that
typed code makes about untyped components, thereby ensuring sound types.

Unfortunately, the high run-time cost of sound types has split
the community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but these systems are all unsound [CITE].
They intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
graduality and its extension to new types and correctness theorems [CITE].
A few academics have studied the cost of run-time checks
in detail [CITE] and proposed implementation methods [CITE],
compiler technology [CITE],
and even weakened semantics [CITE], but none of these efforts
decisively close the performance gap.
The most promising attempt is the \emph{concrete} semantics
for gradual types [CITE kafka nom etc], but its low performance
overhead stems from severe restrictions on untyped code.
These restrictions may not be acceptable to the maintainers of a
working untyped codebase.
Indeed, the maintainers might prefer to re-implement their system
in a typed language as Dropbox recently did to its core sync
engine.\footnote{\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}}

@@

This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a sound type system for a subset of
Python along with a runtime system designed to lower the cost of run-time checks
and capitalize on the benefits of sound types.
The language design is a novel twist on concrete types that lets programmers
begin with a shallow claims about data structures and gradually opt in to deep soundness.
Remarkably, the performance of Python is an upper bound on their observed performance of
Static Python.

@@

We have written this paper to promote two future directions.
First, we want to encourage system-builders to reproduce the essentials of
\SP{} in another context and measure the net performance.
Second, we want to lend focus to theoretical research on gradual typing.
On one hand, the restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
may be worthwhile assumptions for theory going forward.
On the flip side, there are issues in today's \SP{} that might
be addressed by future research.

>>>>

\acknowledgements{
  Thanks to
   Guido van Rossom for stimulating tweets.
  This work was partly supported by the US National Science Foundation.
  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
}

@@

Gradual typing has attracted significant interest as a solution to
the impasse between static and dynamic typing.
The premise is simple: let programmers introduce types in part of a
codebase and while leaving the rest untyped.
Run-time checks can in principle enforce the assumptions that
typed code makes about untyped components, thereby ensuring that
the types are sound no matter how the untyped code behaves.

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but these systems are all unsound [CITE].
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive types [CITE].
A few academics have studied the cost of run-time checks
in detail [CITE] and proposed implementation methods [CITE],
compiler technology [CITE],
and even weakened semantics [CITE], but none of these efforts
decisively close the performance gap.
The most promising attempt is the \emph{concrete} semantics
for gradual types [CITE kafka nom etc], but its low performance
overhead stems from severe restrictions on untyped code.
These restrictions may not be acceptable to the maintainers of a
working untyped codebase.
Indeed, the maintainers might prefer to re-implement their system
in a typed language, as Dropbox recently did with its core sync
engine.\footnote{\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}}

@@

This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a sound type system for a subset of
Python along with a runtime system designed to
capitalize on the benefits of sound types.
The language design is a variation of concrete types that lets programmers
begin with shallow claims about data structures and gradually opt in to deeper soundness.
Remarkably, the performance of Python is an upper bound on the observed performance of
Static Python.

@@

We have written this paper to promote two future directions.
First, we want to encourage system-builders to reproduce the essentials of
\SP{} in another context and measure the net performance.
Second, we want to lend focus to theoretical research on gradual typing.
The restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
may be worthwhile assumptions for theory going forward.
On the flip side, there are issues in today's \SP{} that might
be addressed by future research.
